---
title: "MIC Course"
subtitle: "Practical Data Munging - HIV Example"
author: "Cliburn Chan, Janice McCarthy"
date: "June 9, 2022"
output: html_document
---

# Practical Data Munging


```{r}
library("tidyverse")
library("readxl")
```

## Objective

Suppose we want to visualize the change in deaths of PWH over time for transmission risk factors.

## Get data and try to understand it


```{r}
df<- read_excel('files/deaths-of-persons-diagnosed-with-hiv-aids.xlsx')
```


```{r}
df  %>% dim
```


```{r}
df %>% head(2)
```


```{r}
df %>% tail(2)
```


```{r}
df  %>% summary
```


```{r}
df %>% str
```

## Pull out the transmission categories into separate data frames


```{r}
df  %>% select(Category) %>% distinct
```


```{r}
df %>%
filter(Category == "Transmission Category: Male Adult or Adolescent")  %>% 
select(Year, -Category, "Transmission Category: Male Adult or Adolescent"=Group, Count) -> df_male
```


```{r}
df_male %>% head(3)
```


```{r}
df %>%
filter(Category == "Transmission Category: Female Adult or Adolescent")  %>% 
    select(Year, -Category, "Transmission Category: Female Adult or Adolescent"=Group, Count) -> df_female
```


```{r}
df_female %>% head(3)
```


```{r}
df %>%
   filter(Category == "Transmission Category: Child (<12 Years Old at the End of Year)")  %>% 
   select(Year, -Category, "Transmission Category: Child (<12 Years Old at the End of Year)"=Group, Count) -> df_child
```

## Combine the transmission categories


```{r}
df_child %>% 
     pivot_wider(names_from = "Transmission Category: Child (<12 Years Old at the End of Year)", values_from = Count) %>% 
     mutate(Category="Child") -> df_child_wide
```


```{r}
df_female  %>% 
    pivot_wider(names_from = "Transmission Category: Female Adult or Adolescent", values_from = Count) %>%
    mutate(Category="Female") -> df_female_wide
```


```{r}
df_male  %>% 
   distinct %>%
   pivot_wider(names_from = `Transmission Category: Male Adult or Adolescent`, values_from = Count) %>% 
   mutate(Category="Male") -> df_male_wide
```


```{r}
bind_rows(df_male_wide, df_female_wide, df_male_wide) -> df_transmission
```


```{r}
df_transmission %>% head(3)
```

## Use `pivot_longer` to make the transmission data tidy


```{r}
df_transmission %>% 
  pivot_longer(cols = c(-Year, -Category), names_to = "Transmission", values_to = "Count")
```

There are NA's in some rows, we'll remove those

```{r}
df_transmission %>% 
  pivot_longer(cols = c(-Year, -Category), names_to = "Transmission", values_to = "Count") %>% 
  na.omit(Count) -> df_transmission_tall
```

```{r}
df_transmission_tall  %>% head
```

## Create a summary table


```{r}
df_transmission_tall %>% 
   group_by(Transmission) %>% 
   summarize(Deaths=sum(Count))
```

## Plot example 1


```{r}
df_transmission_tall %>%
   ggplot(aes(x=Transmission, y=Count, fill=Transmission)) +
   facet_grid(Year ~ Category) +
   geom_bar(stat='Identity') +
   theme(axis.text.x=element_blank(),
      axis.ticks.x=element_blank())
```

## Plot example 2


```{r}
df_transmission_tall %>%
    ggplot(aes(x=Year, y=Count, fill=Transmission)) +
    facet_wrap(.~Transmission, ncol=1, scales="free_y") +
    geom_point() +
    geom_smooth() +
    guides(fill="none") -> g
```


```{r, fig.height= 15, fig.width= 5}
suppressWarnings(print(g))
```

## Exercises

**1**. Read the file `files/persons-living-with-hiv-aids.xlsx` into a data frame.


```{r}

```

**2**. How many rows and columns are there?


```{r}

```

**3**. Is the number of Asians with HIV growing over time? Do this using a bar plot.


```{r}

```

**4**. Make a horizontal bar chart of the total number of PWH over all years for each ethnic group. There should be a single Asian group. The bar chart should display the names of each ethnic group on the y-axis, and the number of deaths on the x-axis, where the ethnic groups are ordered from fewest (top) to most (bottom) deaths.


```{r}

```
