---
title: "MIC Course"
subtitle: "Seurat Objects"
author: "Janice M. McCarthy"
date: "June 9, 2022"
output: html_document
---

```{r}
library(tidyverse)
library(Seurat)
library(SeuratObject)
```

# Seurat Object

R has two different types of data structures that it considers to be 'objects'. In most 'object oriented' programming languages, the essence of an object is that it has both data and functions. R handles that a bit differently, but the idea that the data and functions that operate on it are somehow 'together' holds.

The two different types are called S3 and S4. We don't have time for in-depth coverage of these two different types, but note that the Seurat object is an S4 object that also contains some S3 objects and some S4 object.

Both types of objects are really just lists, but with some additional structure. (We don't have time to go into this in detail, but Advanced R by Hadley Wickham is a good source for those interested.) 

The elements of an S4 object are called 'slots' and can be accessed using the '@' operator. S3 elements, just like elements of lists and dataframes, can be accessed with the '$' operator. The following are the slots of a Seurat object (this comes directly from the [SeuratObject documentation:] (https://cran.r-project.org/web/packages/SeuratObject/SeuratObject.pdf)

- *assays* A list of assays for this project

- *meta.data* Contains meta-information about each cell, starting with number of features detected (nFeature) and the original identity class (orig.ident); more information is added using AddMetaData

- *active.assay* Name of the active, or default, assay; settable using DefaultAssay

- *active.ident* The active cluster identity for this Seurat object; settable using Idents
graphs A list of Graph objects
neighbors ...

- *reductions* A list of dimensional reduction objects for this object

- *images* A list of spatial image objects

- *project.name* Name of the project

- *misc* A list of miscellaneous information

- *version* Version of Seurat this object was built under

- *commands* A list of logged commands run on this Seurat object

- *tools* A list of miscellaneous data generated by other tools, should be filled by developers only using Tool<-

## Built-in data set pbmc_small

```{r}
pbmc_small

slotNames(pbmc_small)
```
We can use 'accessor' functions to get parts of the object.
```{r}
#pbmc_small@assays$RNA@counts

rna_assay <- GetAssay(pbmc_small,assay = "RNA")

```
`rna_assay` is itself an S4 object:

```{r}
slotNames(rna_assay)

```

```{r}
GetAssayData(rna_assay) %>% head(3)
```
The above are 'normalized' counts. To get the raw counts, we can access the `counts` slot.

```{r}
rna_assay@counts %>% head(3)
```

We'll move to the SeuratObject tutorial that the authors of the software created:

https://github.com/satijalab/seurat/wiki/
