---
title: "scSorter demonstration"
output: html_notebook
---

```{r prereqs}
knitr::opts_chunk$set(echo = T, message = FALSE, warning = FALSE)
stdt<-date()

# Libraries
library(tidyverse)
library(Seurat)
library(scSorter)

# Paths
wd <- "/hpc/group/chsi-mic-2022"
intermeddir <- file.path(wd, "intermed")
sigfile = "/hpc/group/chsi-mic-2022/data/cellAssignSig.csv"
```

**Goal:** Show how to use the R package scSorter to classify cells into cells types using gene expression and a signature matrix.

**Data source:** Christian et al. 2021 Cell Reports (<https://doi.org/10.1016/j.celrep.2021.109118>)

## Load data

```{r}
seulist <- readRDS(file.path(intermeddir, "seulist-drc.rds"))
```

## Load signature matrix

See the script `make-cellAssignSig.R` for details on how this file was generated.

```{r}
sigobj <- read.csv(file = sigfile)
head(sigobj)
```

Translate marker genes symbols to gene names used in Seurat objects

First, extract gene names from Seurat objects

```{r}
# extract genes from seurat objects
seulist %>%
  map(~rownames(Seurat::GetAssayData(.x))) %>%
  unlist() %>%
  unique() -> seulist.genes
length(seulist.genes) #13430 genes in total
seulist.genes[1:5]
```

If we transform these into uppercase, do they match markers? Of the 39 markers, 27 have matches.

```{r}
data.frame(gene =  seulist.genes, 
           upper.gene = toupper(seulist.genes)) -> gene.indx

# how many marker matches?
length(unique(sigobj$Marker)) #39 markers
sum(unique(sigobj$Marker) %in% gene.indx$gene) # 0 matches
sum(unique(sigobj$Marker) %in% gene.indx$upper.gene) # 27 matches
```

Which markers do not have matches?

```{r}
unique(sigobj$Marker)[!unique(sigobj$Marker) %in% gene.indx$upper.gene]
```

Recode the signature matrix so that the marker names will match the gene names in the seurat objects

```{r}
sigobj %>%
  left_join(gene.indx, by = c("Marker"="upper.gene")) %>%
  dplyr::select('Type'='Group', 'Marker'='gene', Weight, 'orig.Marker'='Marker') %>%
  filter(!is.na(Marker)) -> sigobj.t
```

## Classify cells

Function to ...

1.  Find variable features in Seurat Object
2.  Find genes from the signature matrix that are also variable genes
3.  Extract expression matrix for those genes
4.  Get predicted cell types using expression matrix and signiture matrix
5.  Add cell types as metadata to the seurat object

```{r}
celltypefun <- function(cobj, sigobj.t) {
  
  # Find variable features
  cobj.vf <- FindVariableFeatures(cobj)
  
  #' Use the union of variable genes and marker genes 
  union(
    Seurat::VariableFeatures(cobj.vf), sigobj.t$Marker
  ) -> topgenes
  #topgenes
  
  #' get expression matrix
  dat <- Seurat::GetAssayData(cobj.vf)
  exprmat <- as.matrix(dat[row.names(dat) %in% topgenes,])
  print(dim(exprmat))
  
  #' Get cell type prediction
  print(paste0("scSorter started", Sys.time()))
  scSorter::scSorter(
    expr = exprmat,
    anno = sigobj.t
  ) -> ctpred
  print(paste0("scSorter finished", Sys.time()))
  
  #' Add inferred cell types as meta data
  Seurat::AddMetaData(
    object = cobj.vf,
    metadata = stats::setNames(ctpred$Pred_Type, colnames(cobj.vf)),
    col.name = "scSorter"
  ) -> cobj.vf

  return(cobj.vf)
}
```

Apply the function to each seurat object...

### disTrm

```{r}
seulist[["disTrm"]]
distrm.out <- celltypefun(cobj = seulist[["disTrm"]], 
                          sigobj.t = sigobj.t)

distrm.out@meta.data %>%
  group_by(orig.ident, scSorter) %>%
  summarize(ncells = length(orig.ident))
```

### Tcm

```{r}
seulist[["Tcm"]]
tcm.out <- celltypefun(cobj = seulist[["Tcm"]], 
                          sigobj.t = sigobj.t)
tcm.out@meta.data %>%
  group_by(orig.ident, scSorter) %>%
  summarize(ncells = length(orig.ident))
```

### Tem

```{r}
seulist[["Tem"]]
tem.out <- celltypefun(cobj = seulist[["Tem"]], 
                          sigobj.t = sigobj.t)
tem.out@meta.data %>%
  group_by(orig.ident, scSorter) %>%
  summarize(ncells = length(orig.ident))
```

### Trm

```{r}
seulist[["Trm"]]
trm.out <- celltypefun(cobj = seulist[["Trm"]], 
                          sigobj.t = sigobj.t)
trm.out@meta.data %>%
  group_by(orig.ident, scSorter) %>%
  summarize(ncells = length(orig.ident))
```

### Save

```{r}
names(seulist)
seulist.sc <- list(distrm.out, tcm.out, tem.out, trm.out)
names(seulist.sc) <- names(seulist)
seulist.sc

saveRDS(seulist.sc, file.path(intermeddir, "seulist_sc.rds"))
```
