---
title: "MIC Course"
subtitle: "Ggplot Exercises and Solutions"
author: "Janice M. McCarthy"
date: "June 9, 2022"
output: html_document
---

```{r}
library(tidyverse)
```
# R Graphics

## Exercises

1. Plot a scatter plot of petal width vs petal length from the iris data set. (Recall that in base R, you can access the column for petal width using iris$Petal.Width.)

```{r}
iris %>% 
  ggplot(aes(x=Petal.Width, y = Petal.Length)) +
  geom_point()
```



2. Add a best fit line to the plot.

```{r}
iris %>% 
  ggplot(aes(x=Petal.Width, y = Petal.Length)) +
  geom_point() +
  geom_smooth(method = "lm", level=0.99)
```


3. Create a box plot of petal width by species and another box plot of petal length by species. Make these plots sit side-by-side.

```{r}

iris %>% pivot_longer(cols=c(Petal.Length,Petal.Width),
                      names_to="Vars", values_to = "Inches") %>%
  ggplot(aes(Species, Inches)) + 
  geom_boxplot() + facet_wrap(.~Vars) +
  geom_jitter()
  
```



4. Create a scatter plot of petal width vs length and color/change symbol by species.

```{r}
iris %>% ggplot(aes(Petal.Width, Petal.Length, color = Species)) +
  geom_point()
```


5. Add best fit lines for each species to this plot. 

```{r}
 toSave <- iris %>% ggplot(aes(Petal.Width, Petal.Length, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")

toSave
```

6. Save this scatter plot to  a file called "figs/iris_ggplot.png".

```{r}
ggsave("figs/iris_ggplot.png", toSave )
```


7. Read the plot file back in and display in the markdown.




8. Create a histogram of Petal Length, and fill the bars by Species

```{r}
iris %>% ggplot(aes(x=Petal.Length, fill = Species)) + geom_histogram()
```


9. Create box plots of Sepal Width by Species

```{r}
iris %>% ggplot(aes(x = Sepal.Width, y = Species, color = Species)) + geom_boxplot() +
  geom_jitter()
```

```{r}
iris %>% ggplot(aes(x=Species,y=Petal.Width)) + 
  geom_boxplot() +
  facet_wrap(~Species) 
```


## Great ggplot resource!

http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html